---
title: How the Cubs broke the 108-year drought in 2016
author: Cole Peterson
date: '2022-04-21'
slug: how-the-cubs-broke-the-108-year-drought-in-2016
categories: []
tags: []
---

108 years.

That's how long it had been  since the Chicago Cubs won their last MLB World Series trophy. After over a century, the "Lovable losers" knocked off the then Cleveland Indians to take the series 4-3 and hoist the ultimate prize.

But how did they do it?

Sure they had the NL MVP and Cy Young in Kris Bryant and Jake Arrieta. But baseball is a team sport, so let's take a deeper dive into the Cubs championship run.

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(gt)
library(cowplot)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
offensivestats <- read_csv("Data/MLB Offensive.csv")

fourteams <- c("Chicago Cubs", "Cleveland Indians", "Boston Red Sox", "Los Angeles Dodgers")

references <- offensivestats %>%
  filter(Tm %in% fourteams)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot() + 
  geom_point(data=offensivestats, aes(x=BA, y=OPS, size=BB), color = "black", alpha = .5) +
  geom_point(data=references, aes(x=BA, y=OPS, size=BB), color = "red", alpha = 1) +
  scale_size(range = c(1, 7.5), name="Walks") +
  geom_text(aes(x=.254, y=.7805, label="Chicago Cubs"), color="#0E2240") +
  geom_text(aes(x=.269, y=.758, label="Cleveland Indians"), color="#00471B") +
  geom_text(aes(x=.275, y=.81, label="Boston Red Sox"), color="#ED174C") +
  geom_text(aes(x=.243, y=.72, label="Los Angeles Dodgers"), color="#00538C") +
  labs(
    x="Team Batting Average",
    y="Team On Base Plus Slugging Percentage",
    title="The Cubs Walked Their Way Into A Championship", 
    subtitle="While ranking middle of the pack in team batting average, Chicago found other ways to get it done offensively.", 
    caption="Source: Baseball Reference | By Cole Peterson") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10),
    panel.grid.minor = element_blank(),
    plot.title.position = "plot"
    )
```

Even though the Cubs knew how to get on base, an area of the game that didn't get mentioned as much that season was defense.

Their fielding was on a different level in the 2016 season, as they saved nearly 40 more runs than the next closest team because of their defensive play.

The Cubs defensive efficiency was through the roof, and they knew how to limit scores and make opponents grind for every singles run.

Two Cubs also took home two Golden Glove awards in 2016, Anthony Rizzo and Jason Heyward.

```{r message=FALSE, warning=FALSE, echo=FALSE}
defensivestats <- read_csv("Data/MLB Defense.csv")

ds <- c("Chicago Cubs", "Cleveland Indians", "Los Angeles Dodgers", "Texas Rangers", "Boston Red Sox", "Washington Nationals")

divisionteams <- defensivestats %>%
  filter(Tm %in% ds)

divisiondefensive <- divisionteams %>%
  select(Tm, `RA/G`, `Fld%`, `DefEff▼`, Rtot, Rdrs, DP) %>%
  arrange(`RA/G`)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
divisiondefensive %>% 
  gt() %>% 
  cols_label(
    Tm = "Team",
    `Fld%` = "Fielding Percentage",
    `DefEff▼` ="Defensive Efficiency",
    Rtot = "Runs Players Were Worth On Defense",
    Rdrs = "Defensive Runs Saved Above Average",
    `RA/G` = "Runs Allowed Per Game",
    DP = "Double Plays",
  ) %>%
  tab_header(
    title = "Chicago got it done defensively",
    subtitle = "The Cubs saved more runs than any other team in baseball by nearly 40, while boasting the highest defensive efficiency."
  ) %>% 
  tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) %>% 
  tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) %>%
  tab_source_note(
    source_note = md("**By:** Cole Peterson  |  **Source:** Baseball Reference")
  ) %>%
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
    ) %>%
  opt_row_striping() %>% 
  opt_table_lines("none") %>%
  tab_style(
    style = list(
      cell_fill(color = "#0E3386"),
      cell_text(color = "#CC3433")
      ),
    locations = cells_body(
      rows = Tm == "Chicago Cubs")
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#E31937"),
      cell_text(color = "#0C2340")
      ),
    locations = cells_body(
      rows = Tm == "Cleveland Indians")
  )
```

But how could one forget about pitching.

The Cubs may have had one of the best pitching staffs in the team's long history during their championship run.

Anchored by a rotation of Kyle Hendricks, Jon Lester, John Lackey, and Arrieta, Chicago performed at an elite level on the mound in 2016.

Multiple players ranked in the top 10 in some of the most important pitching categories, namely earned run average and opponent batting average.

They kept the opposition guessing all season en route to a dramatic World Series victory.

```{r message=FALSE, warning=FALSE, echo=FALSE}
mlbadvancedpitching <- read_csv("Data/MLB Advanced Individual Pitching.csv")
regularpitching <- read_csv("Data/MLB PItching 2016.csv") 

mlbpitchingrenamed <- mlbadvancedpitching %>%
  rename(
    Rank=...1,
    Player=...2,
    Age=...3,
    Team=...4,
    BA=...5,
    OBP=...6,
    SLG=...7,
    OPS=...8,
    BAbip=...9,
    `HR%`=...10,
    `SO%`=...11,
    `BB%`=...12,
    EV=...13,
    `HardH%`=...14
  )  %>% mutate(Player = gsub("\\*", "", Player))

statswithoutgarbage <- mlbpitchingrenamed %>%
  filter(Rank != "Rk" & Player != "Name")

statswithnumbers <- statswithoutgarbage %>%
  mutate_at(vars(-Player), as.numeric)

regularpitchingrenamed <-regularpitching %>%
  rename(
    ERA=`ERA▲`,
  ) %>% mutate(Name = gsub("\\*", "", Name))
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
topOppBA <- statswithnumbers %>% top_n(10, wt=-`BA`)

topERA <- regularpitchingrenamed %>% top_n(10, wt=-`ERA`)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
ja <- topOppBA %>% filter(Rank == "10")
kh <- topOppBA %>% filter(Rank == "2")
jl <- topOppBA %>% filter(Rank == "6")
jla <- topOppBA %>% filter(Rank == "13")

kh2 <- topERA %>% filter(Rk == "1")
jl2 <- topERA %>% filter(Rk == "2")
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
bar1 <- ggplot() + 
  geom_bar(data=topOppBA, aes(x=reorder(Player, -`BA`), weight=`BA`)) + 
  geom_bar(data=ja, aes(x=reorder(Player, -`BA`), weight=`BA`), fill="#0E3386") + 
  geom_bar(data=kh, aes(x=reorder(Player, -`BA`), weight=`BA`), fill="#0E3386") + 
  geom_bar(data=jl, aes(x=reorder(Player, -`BA`), weight=`BA`), fill="#0E3386") + 
  geom_bar(data=jla, aes(x=reorder(Player, -`BA`), weight=`BA`), fill="#0E3386") + 
  coord_flip() +
  labs(title = "Arrieta dominated the oppostion in his Cy Young season", subtitle= "Four of the top 10 lowest opponent batting average's belonged to the Cubs pitching staff in 2016.", x="", y="Opponent Batting Average") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 15, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank(),
    plot.title.position = "plot"
    )

bar2 <- ggplot() + 
  geom_bar(data=topERA, aes(x=reorder(Name, -`ERA`), weight=`ERA`)) + 
  geom_bar(data=kh2, aes(x=reorder(Name, -`ERA`), weight=`ERA`), fill="#CC3433") + 
  geom_bar(data=jl2, aes(x=reorder(Name, -`ERA`), weight=`ERA`), fill="#CC3433") + 
  coord_flip() +  
  labs(title = "The Cubs whole rotation also limited earned runs", subtitle= "Chicago had the two lowest earned run averages with Hendricks and Lester dealing in the regular season.", x="", y="Earned Run Average", caption="Source: Baseball Reference | By Cole Peterson") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 15, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank(),
    plot.title.position = "plot"
    )
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
plot_grid(bar1, bar2, ncol=1)
```

It's been long overdue for the city of Chicago to get a World Series trophy, and the 2016 season will go down as one of the most notable MLB teams ever.

They ended the curse of Steve Bartman and stopped the 108-year long drought. The Cubs did it through stellar defense and an elite pitching staff. Even though they did have the NL MVP and one of the highest on-base percentages as a team.

Hopefully it won't be 108 more years till the next banner flies at Wrigley Field.

Note: The code that made these graphics can be found here.